Parsing Postfix log files

### 2009-03-31 doc/logparser-background.tex
Changes from proofreading of introduction chapter.

\begin{table}[ht]
    \caption{Sample output from SQL query showing percentage of delivery
    attempts rejected by each restriction}
    \empty{}\label{Sample output from SQL query}
    \begin{tabular}[]{lrr}
        \tabletopline{}%
        Restriction & Number of  & Percentage of    \\
                    & rejections & total rejections \\
        \tablemiddleline{}%
        \input{build/include-sample-sql-output}
        \tablebottomline{}%
    \end{tabular}

@@

\subsection{Mixing and Matching Postfix Restrictions}

@@

    \item Greylisting~\cite{greylisting} is a technique that temporarily
        rejects a delivery attempt when the triple of \newline{}
        \tab{}\texttt{(sender address, recipient address, remote
        \acronym{IP} address)}\newline{} has not been seen before; on
        second and subsequent delivery attempts from that triple the mail
        will be accepted.  This blocks spam from some senders because
        maintaining a list of failed addresses and retrying after a
        temporary failure is uneconomical for a spam sender, but a
        legitimate mail server must retry deliveries that failed
        temporarily.  Sadly spam senders are using increasingly complex and
        well written programs to distribute spam, frequently using an ISP
        provided \acronym{SMTP} server from a compromised machine on the
        ISP's network.  Greylisting will slowly become less effective as
        spam senders adapt, but it does block a large percentage of spam
        mail at the moment; the most effective restrictions from the
        \numberOFlogFILES{} log files used when generating the results in
        \sectionref{Results} are shown in \tableref{Summary of rejections}.
        The table shows that greylisting is worth using at the moment,
        particularly when you take into account its position as the final
        restriction that a mail must overcome in the configuration used on
        the mail server that generated the log files: on that server
        greylisting only takes effect for mails that have passed all other
        restrictions.  Some problems may be encountered when using
        greylisting: some servers fail to retry after a temporary failure,
        or legitimate mail may be delayed, particularly when coming from a
        pool of servers.

        \begin{table}[ht]
            \caption{Summary of rejections}\label{Summary of rejections}
            \input{build/include-restriction-summary-table}
        \end{table}

>>>>

\begin{table}[ht]
    \caption{Sample output from SQL query, showing the percentage of
    delivery attempts rejected by top ten restrictions}
    \empty{}\label{Sample output from SQL query}
    \begin{tabular}[]{lrr}
        \tabletopline{}%
        Restriction & Number of  & Percentage of    \\
                    & rejections & total rejections \\
        \tablemiddleline{}%
        \input{build/include-sample-sql-output}
        \tablebottomline{}%
    \end{tabular}

@@

\subsection{Mixing And Matching Postfix Restrictions}

@@

    \item Greylisting~\cite{greylisting} is a technique that temporarily
        rejects a delivery attempt when the triple of \newline{}
        \tab{}\texttt{(sender address, recipient address, remote
        \acronym{IP} address)}\newline{} has not been seen before; on
        second and subsequent delivery attempts from that triple the mail
        will be accepted.  This blocks spam from some senders because
        maintaining a list of failed addresses and retrying after a
        temporary failure is uneconomical for a spam sender, but a
        legitimate mail server must retry deliveries that failed
        temporarily.  Sadly spam senders are using increasingly complex and
        well written programs to distribute spam, frequently using an ISP
        provided \acronym{SMTP} server from a compromised machine on the
        ISP's network.  Greylisting will slowly become less effective as
        spam senders adapt, but it does block a large percentage of spam
        mail at the moment; the most effective restrictions from the
        \numberOFlogFILES{} log files used when generating the results in
        \sectionref{Results} are shown in \tableref{Most effective Postfix
        restrictions}.  The table shows that greylisting is worth using at
        the moment, particularly when you take into account its position as
        the final restriction that a mail must overcome in the
        configuration used on the mail server that generated the log files:
        on that server greylisting only takes effect for mails that have
        passed all other restrictions.  Some problems may be encountered
        when using greylisting: some servers fail to retry after a
        temporary failure, or legitimate mail may be delayed, particularly
        when coming from a pool of servers.

        \begin{table}[ht]
            \caption{Most effective Postfix restrictions}
            \empty{}\label{Most effective Postfix restrictions}
            \input{build/include-restriction-summary-table}
        \end{table}

### 2009-03-31 doc/logparser-introduction.tex
Changes from proofreading of introduction chapter.

The architecture and implementation described in this thesis was developed
as the foundation of a larger project to improve anti-spam defences, by
analysing the performance of the set of filters currently in use,
optimising the order and membership of the set based on that analysis, and
developing supplemental filters where deficiencies are identified.  Most
anti-spam techniques are content-based (e.g.~\cite{a-plan-for-spam,
relaxed-online-svms, word-stemming}) and require the mail to be accepted
before determining if it is spam, but rejecting mail during the delivery
attempt is preferable: senders of non-spam mail that is mistakenly rejected
will receive an immediate non-delivery notice; resource usage is reduced on
the accepting mail server, allowing more intensive content-based techniques
to be used on the mail that is accepted; users have less spam mail to wade
through.  Improving the performance of anti-spam techniques that are
applied when mail is being transferred via \acronym{SMTP} is the goal of
this project, by providing a platform for reasoning about anti-spam
filters.

The approach chosen to measure performance is to analyse the log files
produced by the \acronym{SMTP} server in use, Postfix, rather than
modifying its source code to generate statistics: this approach improves
the chances of other Postfix users testing and using the software developed
for this project.  The need arose for a parser capable of dealing with the
great number and variety of log lines produced by Postfix: the parser must
be designed so that adding support for parsing new inputs is a simple task,
because the log lines to be parsed will change over time.  This variety in
log lines occurs for several reasons:

@@

    \item The mail administrator can define custom rejection messages.

@@

Most mail server administrators will have performed some basic processing
of the log files produced by their mail server at one time or another,
whether it was to debug a problem, explain to a user why their mail is
being rejected, or check if new anti-spam measures are working.  The more
adventurous will have generated statistics to show how successful each of
their anti-spam measures has been in the last week, and possibly even
generated some graphs to clearly illustrate these statistics to management
or users.\footnote{This was the first real foray the author, a Systems
Administrator for a network of over 2,000 computers and over 1,800 users,
took into processing Postfix log files.}  Very few will have performed
in-depth parsing and analysis of their log files, where the parsing must
correlate the log lines per-connection or per-queueid rather than
processing log lines independently.  One of the barriers to this kind of
processing is the unstructured nature of Postfix log files, where each log
line was added on an ad hoc basis as a requirement was discovered or new
functionality was added.\footnote{A history of all changes made to Postfix
is distributed with the source code, available from
\urlLastChecked{http://www.postfix.org/}{2009/02/23}.}  Further
complication arises because the set of rejection messages is not fixed, and
there are many ways in which the log lines may differ between servers, even
within the same organisation, where servers may be configured differently,
or even running different versions of Postfix.

@@

    \item Some ignored the majority of log lines, parsing specific log
        lines accurately, but without any provision for parsing new or
        similar log lines.

    \item Others sloppily parsed the majority of log lines, but were
        incapable of distinguishing between log lines of the same category,
        e.g.\ rejecting a mail delivery attempt.

    \item None of the reviewed parsers extracted enough data for the
        purposes of this project.

@@

The only prior published work on the subject of parsing Postfix log files
that the author is aware of is \textit{Log Mail Analyser: Architecture and
Practical Utilizations\/}~\cite{log-mail-analyser}, which aims to extract
data from log files, correlate it, and present it in a form suitable for a
systems administrator to search using the myriad of standard Unix text
processing utilities already available.  A full state of the art review is
provided in \sectionref{state of the art review}, beginning with a review
of \textit{Log Mail Analyser}.

The architecture developed is conceptually simple: provide a few generic
functions (\textit{actions\/}), each capable of dealing with an entire
category of inputs (e.g.\ rejecting a mail delivery attempt), accompanied
by a multitude of precise patterns (\textit{rules\/}), each of which
recognises one input variant and only that variant (e.g.\ rejection
by a specific \acronym{DNSBL}).  This architecture is ideally suited to
parsing inputs where the input is not fully understood or does not conform
to a fixed grammar: the architecture warns about unparsed inputs and other
errors, but continues parsing as best it can, allowing the developer of a
new parser to decide which deficiencies are most important and require
immediate attention, rather than being forced to fix the first error that
arises.

This thesis documents the difficult process of parsing Postfix log files,
presenting an architecture designed to enable the users of a parser to
easily extend it to parse their particular inputs, without requiring a lot
of work or a high level of understanding of the parsing process and the
parser's internal workings.  This architecture is the basis of
\parsername{}, a program that parses Postfix log files and places the
resulting data into a database for later analysis.  The gathered data can
be used to optimise current anti-spam measures, provide a baseline to test
new anti-spam measures against, or to produce statistics showing how
effective those measures are.  Numerous other uses are possible for such
data: improving server performance by identifying troublesome destinations
and reconfiguring appropriately; identifying regular high volume uses
(e.g.\ customer newsletters) and restricting those uses to off-peak times,
or providing a dedicated service for them; detecting virus outbreaks that
propagate via mail; as a base for billing customers on a shared server.
Preserving the raw data enables users to develop a multitude of uses far
beyond those conceived of by the author.

@@

Chapter~\ref{background} provides background information useful in
understanding the motivation behind the project, \acronym{SMTP}, and
Postfix.

Chapter~\ref{state of the art review} reviews both the previously published
research in this area and other Postfix log file parsers, discussing why
they were deemed unsuitable for the task, including why they could not be
improved or expanded upon.

Chapter~\ref{parser architecture} describes the parser architecture
developed for this project, beginning with an overview, then describing
each of the components of the architecture in detail.  This chapter
concentrates on the abstract, theoretical, implementation-independent
aspects of the architecture; the practical aspects are deferred until
chapter~\ref{Postfix Parser Implementation}.

Chapter~\ref{Postfix Parser Implementation} documents \parsername{}, the
parser based on the architecture described in chapter~\ref{parser
architecture}.  The practical difficulties of implementing each of the
components of the architecture are described, accompanied by the many
complications encountered when parsing Postfix log files, and other
important details of the implementation.

Chapter~\ref{Results} evaluates \parsernames{} efficiency, discussing the
various optimisations implemented in the framework, and the effect they
have.  It also discusses the coverage achieved by \parsername{} over
\numberOFlogFILES{} log files, with separate discussions of the number of
log lines correctly recognised, and the number of connections and mails
whose journey through Postfix was correctly reconstructed.

>>>>

The architecture and implementation described in this thesis were developed
as the foundation of a larger project to improve anti-spam defences, by
analysing the performance of the set of anti-spam techniques currently in
use, optimising the order and membership of the set based on that analysis,
and developing supplemental anti-spam techniques where deficiencies are
identified.  Most anti-spam techniques are content-based
(e.g.~\cite{a-plan-for-spam, relaxed-online-svms, word-stemming}) and
require a mail to be accepted before determining if it is spam, but
rejecting mail during the delivery attempt is preferable: senders of
non-spam mail that is mistakenly rejected will receive an immediate
non-delivery notice; resource usage is reduced on the accepting mail
server, allowing more intensive content-based techniques to be used on the
remaining mail that is accepted; users have less spam mail to wade through.
Improving the performance of anti-spam techniques that are applied when
mail is being transferred via \acronym{SMTP} is the goal of this project,
by providing a platform for reasoning about the performance of anti-spam
techniques.

The approach chosen to measure performance is to analyse the log files
produced by the \acronym{SMTP} server in use, Postfix, rather than
modifying its source code to generate statistics: this approach improves
the chances of other Postfix users testing and using the software developed
for this project.  The need arose for a parser capable of dealing with the
great number and variety of log lines produced by Postfix: the parser must
be designed and implemented so that adding support for parsing new inputs
is a simple task, because the log lines to be parsed will change over time.
This variety in log lines occurs for several reasons:

@@

    \item The mail server administrator can define custom rejection
        messages.

@@

Most mail server administrators will have performed some basic processing
of the log files produced by their mail server at one time or another,
whether it was to debug a problem, explain to a user why their mail is
being rejected, or check if new anti-spam techniques are working.  The more
adventurous will have generated statistics to show how successful each of
their anti-spam measures has been in the last week, and possibly even
generated some graphs to clearly illustrate these statistics to management
or users.\footnote{This was the first real foray the author, a Systems
Administrator for a network of over 2000 computers and over 1800 users,
took into processing Postfix log files.}  Very few will have performed
in-depth parsing and analysis of their log files, where the parsing must
correlate the log lines per-connection or per-queueid rather than
processing log lines independently.  One of the barriers to this kind of
processing is the unstructured nature of Postfix log files, where each log
line was added on an ad hoc basis as a requirement was discovered or new
functionality was added.\footnote{A history of all changes made to Postfix
is distributed with the source code, available from
\urlLastChecked{http://www.postfix.org/}{2009/02/23}.}  Further
complication arises because the set of log lines is not fixed, and there
are many ways in which log lines may differ between servers, even within
the same organisation, where servers may be configured differently or even
running different versions of Postfix.

@@

    \item Some parsers ignored the majority of log lines, parsing specific
        log lines accurately, but without any provision for parsing new or
        similar log lines.

    \item Other parsers sloppily parsed the majority of log lines, but were
        incapable of distinguishing between log lines in one same category,
        e.g.\ not distinguishing between different anti-spam techniques
        causing the rejection of delivery attempts.

    \item None of the reviewed parsers extracted enough data for future
        in-depth analysis.

@@

The only prior published work on the subject of parsing Postfix log files
that the author is aware of is \textit{Log Mail Analyser: Architecture and
Practical Utilizations\/}~\cite{log-mail-analyser}, which aims to extract
data from log files, correlate it, and present it in a form suitable for a
systems administrator to search using the myriad of standard Unix text
processing utilities already available; it is reviewed alongside the other
parsers in the State of the Art Review in chapter~\ref{state of the art
review}.

Once it was decided that a new parser would be written, an architecture was
required to base the implementation on.  Existing architectures are
tailored towards parsing inputs with a fixed grammar or a tightly
constrained format, whereas Postfix log files lack any form of constraint,
as outlined earlier.  A new architecture was designed and developed for
this parser, with the hope that it will be useful to others.  The resulting
architecture is conceptually simple: provide a few generic functions
(\textit{actions\/}), each capable of dealing with an entire category of
inputs (e.g.\ rejecting a mail delivery attempt), accompanied by a
multitude of precise patterns (\textit{rules\/}), each of which recognises
one input variant and only that variant (e.g.\ rejection by a specific
\acronym{DNSBL}), and specifies which action will process the inputs it
recognises.  This architecture is ideally suited to parsing inputs
where the input is not fully understood or does not conform to a fixed
grammar: the architecture warns about unparsed inputs and other errors, but
continues parsing as best it can, allowing the developer of a new parser to
decide which deficiencies are most important and require immediate
attention, rather than being forced to fix the first error that arises.

This thesis documents the difficult process of parsing Postfix log files,
presenting an architecture designed to enable the users of a parser to
easily extend it to parse their particular inputs, without requiring a lot
of work or a high level of understanding of the parsing process and the
parser's internal workings.  This architecture is the basis of
\parsername{}, a program that parses Postfix log files and places the
resulting data into a database for later analysis.  The gathered data can
be used to optimise current anti-spam defences, to produce statistics
showing how effective each technique in use is, or to provide a baseline to
test new anti-spam measures against.  Numerous other uses are possible for
such data: improving server performance by identifying troublesome
destinations and reconfiguring appropriately; identifying regular high
volume uses (e.g.\ customer newsletters) and restricting those uses to
off-peak times, or providing a dedicated service for them; detecting virus
outbreaks that propagate via mail; as a base for billing customers on a
shared server.  Preserving the raw data enables users to develop a
multitude of uses far beyond those conceived of by the author.

@@

Chapter~\ref{background} provides background information useful in
understanding \acronym{SMTP}, Postfix, and the motivation behind the
project.

Chapter~\ref{state of the art review} reviews the previously published
research in this area and other Postfix log file parsers, discussing why
they were deemed unsuitable for the task, including why they could not be
improved or expanded upon.

Chapter~\ref{parser architecture} describes the parser architecture
developed for this project, beginning with an overview, then describing
each of the components of the architecture in detail.  This chapter
concentrates on the abstract, theoretical, implementation-independent
aspects of the architecture; discussion of the practical aspects is
deferred until chapter~\ref{Postfix Parser Implementation}.

Chapter~\ref{Postfix Parser Implementation} documents \parsername{}, the
parser based on the architecture described in chapter~\ref{parser
architecture}.  The practical difficulties of implementing each of the
components of the architecture are described, accompanied by the many
complications encountered when parsing Postfix log files, and other details
of the implementation.

Chapter~\ref{Results} evaluates \parsernames{} efficiency, discussing the
various optimisations implemented in the framework, and the effect they
have.  It also discusses the coverage achieved by \parsername{} over
\numberOFlogFILES{} log files, with separate sections for the number of log
lines correctly recognised, and the number of connections and mails whose
journey through Postfix was correctly reconstructed.

### 2009-03-31 doc/logparser-results.tex
Major work on the introduction chapter: the blurb at the start has been extended with content from the SGAI paper, and the layout section is more or less finished.
Changes from proofreading of introduction chapter.

\chapter{Results and Evaluation}

@@

When generating the timing data used in this section, \numberOFlogFILES{}
log files (totaling 10.08 GB, \numberOFlogLINEShuman{} log lines) were each
parsed 10 times, and the parsing times from the 10 runs averaged.  Saving
results to the database was disabled for the test runs, because that
dominates the run time of the program, and the tests are aimed at measuring
the speed of \parsername{} rather than the speed of the database and the
disks the database is stored on.  The computer used for test runs was a
Dell Optiplex 745, shown in \tableref{Computer used to generate
statistics}.

\begin{table}[htbp]
    \caption{Computer used to generate statistics}
    \empty{}\label{Computer used to generate statistics}
    \begin{tabular}[]{ll}
        \tabletopline{}%
        Component  & Component in use                                   \\
        \tablemiddleline{}%
        CPU         & One dual core 2.40GHz Intel\textregistered{}
                        Core\texttrademark{}2 CPU,                      \\
                    & with 32KB L1 cache and 4MB L2 cache.              \\
        RAM         & 2GB 667 MHz DDR RAM\@.                            \\
        Hard disk   & One Seagate Barracuda 7200 RPM 250GB SATA disk.   \\
        \tablebottomline{}%
    \end{tabular}
\end{table}

@@

\showgraph{build/graph-input-size-vs-parsing-time}{Parsing time, file size,
and number of log lines}{parsing time vs file size vs number of log lines
graph}

@@

\subsection{Explaining the Peaks in Log File Size}

@@

\subsection{Rule Ordering For Efficiency}

@@

\showgraph{build/graph-hits}{Hits per rule}{rule hits graph}

>>>>

\chapter{Results And Evaluation}

@@

When generating the timing data used in this section, \numberOFlogFILES{}
log files (totaling 10.08 GB, \numberOFlogLINEShuman{} log lines) were each
parsed 10 times, and the parsing times from the 10 runs averaged.  Saving
results to the database was disabled for the test runs, because that
dominates the run time of the program, and the tests are aimed at measuring
the speed of \parsername{} rather than the speed of the database and the
disks the database is stored on.  The computer used for test runs was a
Dell Optiplex 745, shown in \tableref{Details of the computer used to
generate statistics}.

\begin{table}[htbp]
    \caption{Details of the computer used to generate statistics}
    \empty{}\label{Details of the computer used to generate statistics}
    \begin{tabular}[]{ll}
        \tabletopline{}%
        Component  & Component in use                                   \\
        \tablemiddleline{}%
        CPU         & One dual core 2.40GHz Intel\textregistered{}
                        Core\texttrademark{}2 CPU,                      \\
                    & with 32KB L1 cache and 4MB L2 cache.              \\
        RAM         & 2GB 667 MHz DDR RAM\@.                            \\
        Hard disk   & One Seagate Barracuda 7200 RPM 250GB SATA disk.   \\
        \tablebottomline{}%
    \end{tabular}
\end{table}

@@

\showgraph{build/graph-input-size-vs-parsing-time}{Parsing time, file size,
and number of log lines for \numberOFlogFILES{} log files}{parsing time vs
file size vs number of log lines graph}

@@

\subsection{Explaining The Peaks In Log File Size}

@@

\clearpage{} % XXX GRAPHS MOVE AROUND A LOT
\subsection{Rule Ordering For Efficiency}

@@

\clearpage{} % XXX GRAPHS MOVE AROUND DURING REBUILDS

\showgraph{build/graph-hits}{Log lines recognised per rule}{rule hits graph}

### 2009-03-31 doc/logparser.tex
Changes from proofreading of introduction chapter.

% The name of the program, so I only have to change it in one place.
\newcommand{\parsername}{\acronym{PLP}}
\newcommand{\parsernames}{\acronym['s]{PLP}}
\newcommand{\parsernamelong}{Postfix Log Parser}

@@

    Parsing Postfix log files is much more difficult than it first appears,
    but it \textit{is\/} possible to achieve a high degree of accuracy in
    understanding the log files, and thus accuracy in reconstructing the
    actions taken by Postfix to generate the log files.  This paper
    describes the creation of a parser for Postfix log files, documenting
    the architecture developed and the parser implementing it, explaining
    the difficulties encountered and the solutions developed, with
    reference to an implementation that stores data gleaned from the log
    files in an SQL database.  The gathered data can then be used to
    optimise current anti-spam measures, provide a baseline to test new
    anti-spam measures against, or to produce statistics showing how
    effective those measures are.

>>>>

% The name of the program, so I only have to change it in one place.
\newcommand{\parsername}{\acronym{PLP}}
\newcommand{\parsernames}{\acronym['s]{PLP}}
\newcommand{\parsernamelong}{Postfix Log Parser}
\newcommand{\parsernameshort}{PLP}

@@

    Parsing Postfix log files is much more difficult than it first appears,
    but it \textit{is\/} possible to achieve a high degree of accuracy in
    understanding the log files and reconstructing the actions taken by
    Postfix to generate the log files.  This paper describes the creation
    of a parser for Postfix log files, documenting the architecture
    developed and the parser implementing it, explaining the difficulties
    encountered and the solutions developed, with reference to an
    implementation that stores data gleaned from the log files in an SQL
    database.  The gathered data can then be used to optimise current
    anti-spam measures, provide a baseline to test new anti-spam measures
    against, or to produce statistics showing how effective those measures
    are.
